Type theory in type theory with single substitutions

### 2024-09-17 types_2024_postproc/p.tex
stuff

@@

There are equations in B-systems that cannot be derived from the equations of
SSC. The following four equations are such, they are versions of equations in
section \ref{sec:ssc} lifted under arbitrary amount of $\blank^+$s.
\begin{equation}\label{eq:lifted}
\begin{aligned}
  B[\p^{+^n}][(\gamma^+)^{+^n}] & = B[\gamma^{+^n}][\p^{+^n}] \\
  B[\p^{+^n}][\langle a \rangle^{+^n}] & = B \\
  B[\langle a \rangle^{+^n}][\gamma^{+^n}] & = B[(\gamma^+)^{+^n}][\langle a[\gamma] \rangle^{+^n}] \\
  B[(\p^+)^{+^n}][\langle\q\rangle^{+^n}] & = B
\end{aligned}
\end{equation}
In addition, we need the corresponding lifted equations for terms. Note that we
do not have the non-lifted term versions of the last two equations in SSC.

@@

We define $\alpha$-normal forms as inductive predicates in Prop

separate weakening and single substitution

Define instantiation operations by recursion on $\alpha$-normal forms

Instead of doing induction on the syntax for each of the equations in
(\ref{eq:lifted}), we define a general lemma which can lift any equation between
instantiations over a telescope.

We define $\Sub^*$ to be $\Sub$ with freely added identity and composition
operations. All instantiation operations, lifting operations, and substitution
rules can be redefined for $\Sub^*$.

\begin{lemma}
  Given and $\sigma, \delta : \Sub^*\,\Delta\,\Gamma$, if $A[\sigma] =
  A[\delta]$ and $x[\sigma] = x[\delta]$ for any $A:\Ty\,\Gamma$ and variable
  $x$ in $A$, then
  \begin{itemize}
    \item $\Omega[\sigma] = \Omega[\delta]$ for $\Omega:\Tel\,\Gamma$
    \item $A[\sigma^{+^\Omega}] = A[\delta^{+^\Omega}]$ for $A:\Ty\,(\Gamma+\Omega)$
    \item $a[\sigma^{+^\Omega}] = a[\delta^{+^\Omega}]$ for $a:\Tm\,(\Gamma+\Omega)\,A$
  \end{itemize}
\end{lemma}
\begin{proof}
  asd
\end{proof}

We can simulate parallel substitutions using $\Sub^*$ as iterated single
substitutions, however it does not satisfy the equations of parallel
substitutions. Thus we define parallel substitutions as a list of terms with a
map into $\Sub^*$, and reuse instantiation operation of $\Sub^*$ to convert it
into a sequence of instantiations of single substitutions.
\begin{alignat*}{10}
  & \Tms && : \Con\ra\Con\ra\Con \\
  & \llcorner\blank\lrcorner && : \Tms\,\Delta\,\Gamma \to \Sub^*\,\Delta\,\Gamma \\
  & \upvarepsilon && : \Tms\,\Gamma\,\diamond \\
  & \blank,\blank && : (\gamma:\Tms\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\llcorner\gamma\lrcorner])\ra\Tms\,\Delta\,(\Gamma\ext A) \\
  & \llcorner\upvarepsilon\lrcorner && := id\circ\dots\circ\p \\
  & \llcorner\gamma,a\lrcorner && := \llcorner\gamma\lrcorner^+\circ\langle a \rangle
\end{alignat*}

@@

@@

>>>>

We mainly work in ETT.

@@

The following four equations correspond to equations in B-systems which cannot
be derived from the equations of SSC, these are versions of equations in section
\ref{sec:tt} lifted under arbitrary amount of $\blank^+$s.
\begin{equation}\label{eq:lifted}
\begin{aligned}
  B[\p^{+^n}][(\gamma^+)^{+^n}] & = B[\gamma^{+^n}][\p^{+^n}] \\
  B[\p^{+^n}][\langle a \rangle^{+^n}] & = B \\
  B[\langle a \rangle^{+^n}][\gamma^{+^n}] & = B[(\gamma^+)^{+^n}][\langle a[\gamma] \rangle^{+^n}] \\
  B[(\p^+)^{+^n}][\langle\q\rangle^{+^n}] & = B
\end{aligned}
\end{equation}
The correspoinding equations on terms are also needed. Note that we do not have
the non-lifted term versions of the last two equations in SSC.

@@

A type or term is $\alpha$-normal if it does not contain instantiation
operations, except at the leaves of the syntax tree as variables. However,
$\alpha$-normal can still contain $\beta$/$\eta$ redexes, so that we do not need
to do full normalization. We define variables and $\alpha$-normal forms as
inductive predicates in Prop.
\begin{itemize}
  \item $\q:\Tm\,(\Gamma\ext A)\ (A[\p])$ is a variable.
  \item $x[\p]:\Tm\,(\Gamma\ext A)\ (B[\p])$ is a variable if $x:\Tm\,\Gamma\,B$
  is a variable.
  \item $x:\Tm\,\Gamma\,A$ is $\alpha$-normal if $x$ is a variable.
  \item $\Pi\,A\,B:\Ty\,\Gamma\,i$ is $\alpha$-normal if $A$ and $B$ are
  $\alpha$-normal types.
  \item $\app\,f\,a:\Tm\,\Gamma\,(B[\langle a \rangle])$ is $\alpha$-normal if
  $A$ and $B$ are $\alpha$-normal types, $f:\Tm\,\Gamma\,(\Pi\,A\,B)$ and
  $a:\Tm\,\Gamma\,A$ are $\alpha$-normal terms.
\end{itemize}
The predicate is defined similarly for the other type and term formers. Notably,
we do not state that instantiated types/terms are $\alpha$-normal. It is
important to truncate the $\alpha$-normal predicates to be propositions to
ensure that it preserves the $\beta$/$\eta$ conversion rules.

We define instantiation operations by recursion on $\alpha$-normal forms. The
instantiation operations need to be defined separately for weakenings and single
substitutions for the recursion to be well-founded.

$x[\p][\gamma^+]' = b[\gamma]'[\p]'$

\begin{itemize}
  \item $\p$ is a weakening.
  \item $\gamma^+$ is a weakening if $\gamma$ is a weakening.
  \item $\langle a \rangle$ is an $\alpha$-normal single substitution if $a$ is an
  $\alpha$-normal term.
  \item $\gamma^+$ is an $\alpha$-normal single substitution if $\gamma$ is an
  $\alpha$-normal single substitution.
\end{itemize}

\begin{lemma}
  \label{thm:alpha}
  Every type and term is/has? an $\alpha$-normal form.
\end{lemma}
\begin{proof}
  By induction on the syntax, normalizing the substitutions at the same time.
\end{proof}

Instead of doing induction on the syntax for each of the equations in
(\ref{eq:lifted}), we define a general lemma which can lift any equation between
instantiations over a telescope. For this we define $\Sub^*$ to be $\Sub$ with
freely added identity and composition operations, we do not require it to
satisfy the laws of a category. All instantiation operations, lifting
operations, and substitution rules can be redefined for $\Sub^*$.
\begin{lemma}
  \label{thm:lifting}
  Given and $\sigma, \delta : \Sub^*\,\Delta\,\Gamma$, if $A[\sigma] =
  A[\delta]$ and $x[\sigma] = x[\delta]$ for any $A:\Ty\,\Gamma$ and variable
  $x:\Tm\,\Gamma\,A$, then
  \begin{itemize}
    \item $\Omega[\sigma] = \Omega[\delta]$ for $\Omega:\Tel\,\Gamma$
    \item $A[\sigma^{+^\Omega}] = A[\delta^{+^\Omega}]$ for $A:\Ty\,(\Gamma+\Omega)$
    \item $a[\sigma^{+^\Omega}] = a[\delta^{+^\Omega}]$ for $a:\Tm\,(\Gamma+\Omega)\,A$
  \end{itemize}
\end{lemma}
\begin{proof}
  Assuming the first equation, we prove that $x[\sigma^{+^\Omega}] =
  x[\delta^{+^\Omega}]$ for any variable $x$ by induction on the telescope and
  the variable. Then the two latter equations can be proven by induction on
  $\alpha$-normalized types and terms, still assuming the first equation.
  Finally we prove the first equation by induction on the telescope.
\end{proof}

We can simulate parallel substitutions using $\Sub^*$ as iterated single
substitutions, however it does not satisfy the equations of parallel
substitutions. Thus we define parallel substitutions as a list of terms with a
map into $\Sub^*$, and reuse instantiation operation of $\Sub^*$ to convert it
into a sequence of instantiations of single substitutions.
\begin{alignat*}{10}
  & \Tms && : \Con\ra\Con\ra\Con \\
  & \llcorner\blank\lrcorner && : \Tms\,\Delta\,\Gamma \to \Sub^*\,\Delta\,\Gamma \\
  & \epsilon && : \Tms\,\Gamma\,\diamond \\
  & \blank,\blank && : (\gamma:\Tms\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\llcorner\gamma\lrcorner])\ra\Tms\,\Delta\,(\Gamma\ext A) \\
  & \llcorner\epsilon\lrcorner && := id\circ\p\circ\dots\circ\p \\
  & \llcorner\gamma,a\lrcorner && := \llcorner\gamma\lrcorner^+\circ\langle a \rangle
\end{alignat*}
The computation of instantiating with $\Tms$ is illustrated bellow.
\begin{gather*}
  B[\llcorner\epsilon,a_1,a_2,\dots,a_{n-1},a_n\lrcorner] = \\
  B[\p^{+^n}]\dots[\p^{+^n}][\langle a_1 \rangle^{+^{n-1}}][\langle a_2 \rangle^{+^{n-2}}]\dots[\langle a_{n-1} \rangle^+][\langle a_n \rangle]
\end{gather*}

@@

contextual isomorphism \cite{DBLP:journals/corr/abs-2211-07487}

@@

If we have var, then we cannot do termification to get a CwF from a var-SSC.
